A Graph Theory Based Approach to MDAO Problem Formulation

### 2013-04-19 example_problem.tex
Dr. German's edits

\section{Example Problem}
	\label{s:example problem}
	This section presents an example problem to demonstrate the process of obtaining the FPG and the advantages of doing so. The algorithm from Sec.~\ref{s:building graphs}.\ref{ss:obtaining FPG} was implemented in the NetworkX package of the Python programming language. Any language with the ability to manipulate a graph should be applicable. NetworkX was chosen due to the built--in features, such as cycle detection.

	The example task is the conceptual sizing of a single--aisle subsonic transport using an MDAO framework with objectives for performance and total weight for a required mission. The full set of analysis codes available for use is given in Table \ref{t:analysis codes}.
	\begin{table}[htbp]
	  \centering
	  \caption{Analysis codes for subsonic transport sizing}
		\begin{tabular}{cc}
		\toprule
		analysis code & description \\
		\midrule
		VSP   & parametric geometry \\
		PDCYL & wing and fuselage weight estimation \\
		NPSS  & engine sizing and performance \\
		VORLAX & aerodynamics using the vortex lattice method \\
		PMARC & aerodynamics using a low-order panel method \\
		WATE  & engine weight estimation \\
		FLOPSa & mission performance, \\
		  & engine sizing, and weight estimation \\
		FLOPSb & mission performance only \\
		\bottomrule
		\end{tabular}%
	  \label{t:analysis codes}%
	\end{table}%
	Each of these analysis codes contribute individual disciplinary analysis but they are not mutually exclusive. For example, VORLAX and PMARC are both aerodynamics codes that predict inviscid drag but with different levels of fidelity. The Flight Optimization System (FLOPS) is included twice to represent two different uses. FLOPSa denotes FLOPS being executed for both mission performance and engine analysis, while FLOPSb indicates that FLOPS is executed for only mission performance analysis. This sort of representation is useful for ``supercodes'' that care capable of being used in different ways and with different sets of inputs and outputs.

	In this example, the MCG is formed using a consistant variable naming convention and then connecting every variable with the same name, though this is not the only way to form an MCG. 
	Table \ref{t:ins and outs} presents the full list of variables in the leftmost column and then indicates whether the variable is an input or an output of each analysis code. 
	Some of the variables, such as geometry and performance, represent many different variables which have been bundled together due to there similarity. 
	This bundling is not necessary and does not limit the generality of this example, it is done to simplify the presentation.
	\begin{table}[htb!]
	  \centering
	  \caption{Analysis code input and output description}
		\begin{tabular}{ccccccccc}
		\toprule
		variable & \multicolumn{8}{c}{analysis code} \\
		\midrule
			  & VSP   & PDCYL & NPSS  & VORLAX & PMARC & WATE  & FLOPSa & FLOPSb \\
		geometry & in    & in    &       & in    & in    & in    & in    & in \\
		number of engines &       &       & in    &       &       &       & in    & in \\
		mission &       &       &       &       &       &       & in    & in \\
		fuselage weight &       & out   &       &       &       &       & in    & in \\
		wing weight &       & out   &       &       &       &       & in    & in \\
		engine weight &       &       &       &       &       & out   & out   & in \\
		wetted area & out   &       &       &       &       &       & in    & in \\
		inviscid drag &       &       &       & out   & out   &       &       & in \\
		drag  &       &       & in    &       &       &       & out   & out \\
		engine performance &       &       & out   &       &       & in    & out   & in \\
		performance &       &       &       &       &       &       & out   & out \\
		total weight &       & in    &       &       &       &       & out   & out \\
		\bottomrule
		\end{tabular}
	  \label{t:ins and outs}
	\end{table}

	The MCG $M$ is formed in four steps:
	\begin{enumerate}
	\item An analysis block is created for each analysis code using the information in each column of Table \ref{t:ins and outs}. Each analysis block is formed by creating variable nodes for each input and directing them into a single model node, which is directed into another model node, which is then directed into variable nodes corresponding to each output. 
	A sample analysis block is shown for analysis code FLOPSb in Fig.~\ref{f:FLOPSb analysis block}.
	\begin{figure}[htb!]
	  \begin{center}
		\includegraphics[width=.6\textwidth]{images/FLOPSb_analysis_block}
	  \end{center}
	  \caption{Sample analysis block for analysis code FLOPSb using Table \ref{t:ins and outs}}
	\label{f:FLOPSb analysis block}
	\end{figure}

	\item Variable ((change)) nodes are created to represent the objectives for performance and total weight.

@@

	\item Explicit edges are created from each variable node to every other variable node representing variables with the same name. The direction is determined by whether the variable node has an edge directed into or out of a model node (local input or local output).
	\end{enumerate}
	The resulting MCG is shown in Fig.~\ref{f:design vars}. Emerging cycles are shown in yellow and indicate potential cycles that may exist when an FPG is formed.
	\begin{figure}[htb!]
	  \begin{center}
		\includegraphics[width=.6\textwidth]{images/MCG}
	  \end{center}
	  \caption{Maximal connectivity graph for the subsonic transport example problem}
	\end{figure}

@@

	As presented in Sec.~\ref{s:building graphs}.\ref{ss:obtaining FPG}, the first step to obtaining an FPG is to detect holes. To start, we set the lower indegree limit for every node to be unity to find any potential holes. In this case, it turns out that the variable nodes representing the number of engines is a hole for NPSS, FLOPSa, and FLOPSb. This implies than an FPG cannot be obtained from the given MCG, as shown in Fig.~\ref{f:MCG nengines}. In this figure, the green nodes indicate analysis blocks with a hole, and the cyan nodes indicate analysis blocks that became unused after the invalid analyses were removed ((connections were removed, color edges too)).
	\begin{figure}[htb!]
	  \begin{center}
		\includegraphics[width=.6\textwidth]{images/MCG_nengines}
	  \end{center}
	  \caption{The MCG with one global variable missing. The resulting holes are shown in green and unused analyses shown in cyan.}
	\label{f:MCG nengines}
	\end{figure}

	Therefore, we may now create another variable node to represent the number of engines as a global input and also create explicit edges to supply this new input. Now this step is repeated with the new MCG and no holes are found.

	While Sec.~\ref{s:building graphs}.\ref{ss:obtaining FPG} indicated the process for obtaining an FPG, it did not specify how to decide which edges to keep when resolving a collision, that freedom is left to the specific implementation. The benefit of this approach is that standard graph theory algorithms can be used to make these decisions. The first example of this is to obtain an FPG with the fewest possible number of cycles, which are found using the implementation of Johnson's algorithm \cite{Johnson1975} in the Python package NetworkX. The FPG with the fewest cycles is shown in Fig.~\ref{f:FPG fewest cycles}, which indicates one cycle shown in yellow. A similar alternative would be to minimize the number of analysis blocks involved in cycles, counting multiplicity.
	\begin{figure}[htb!]
	  \begin{center}
		\includegraphics[width=.6\textwidth]{images/FPG_fewest_cycles}
	  \end{center}
	  \caption{FPG with the fewest number of cycles.}
	\label{f:FPG fewest cycles}
	\end{figure}

	Alternatively, it may be desirable to resolving collisions by including the higher fidelity codes, such as NPSS in this case.
	One method to ensure that the edges directed from certain analysis blocks are the ones chosen to resolve the conflicts is to use a ranking system. For this case, consider the rankings for each analysis code given in Table \ref{t:rankings}.
	\begin{table}[htbp]
	  \centering
	  \caption{Example ranking of importance}
		\begin{tabular}{cc}
		\toprule
		analysis code & rank \\
		\midrule
		VSP   & 5 \\
		PDCYL & 5 \\
		NPSS  & 4 \\
		PMARC & 4 \\
		FLOPSb & 4 \\
		VORLAX & 3 \\
		WATE  & 3 \\
		FLOPSa & 2 \\
		\bottomrule
		\end{tabular}%
	  \label{t:rankings}%
	\end{table}% 
	Step two in Sec.~\ref{s:building graphs}.\ref{ss:obtaining FPG} (the resolution of collisions) is now conducted by selecting the edges directed from the analysis blocks with the highest rank for each collision. The resulting FPG is shown in Fig.~\ref{f:FPG highest rank}, and this graph has four cycles, which are enumerated in Table \ref{t:cycles}.
	\begin{figure}[htb!]
	  \begin{center}
		\includegraphics[width=.6\textwidth]{images/FPG_highest_rank}
	  \end{center}
	  \caption{FPG obtained using the ranking system.}
	\label{f:FPG highest rank}
	\end{figure}
	\begin{table}[htbp]
	  \centering
	  \caption{Cycles for the FPG obtained using ranking}
		\begin{tabular}{ccccccc}
		\toprule
		output/input & analysis code & output/input & analysis code & output/input & analysis code & output/input \\
		\midrule
		engine weight & FLOPSb & drag  & NPSS  & engine performance & WATE  & engine weight \\
		engine performance & FLOPSb & drag  & NPSS  & engine performance &       &  \\
		fuselage weight & FLOPSb & total weight & PDCYL & fuselage weight &       &  \\
		wing weight & FLOBSb & total weight & PDCYL & wing weight &       &  \\
		\bottomrule
		\end{tabular}%
	  \label{t:cycles}
	\end{table}%

	Finally, now consider the case where it is desired to have the inviscid drag input into FLOPSb be a multi--fidelity input, meaning that it is sought to use multiple analysis codes to calculate the same variable. This is done simply by setting the upper indegree limit for this node as two and then repreating the (automated) process. In this case, both VORLAX and PMARC are retained, resulting in an FPG that omits only FLOPSa.

%\begin{figure}[htb!]
%  \begin{center}
%    \includegraphics[width=.6\textwidth]{images/FPG_highest_rank_simplified}
%  \end{center}
%  \caption{Simple notional problem with two potential design variables \label{f:design vars}}
%\end{figure}

>>>>

\section{Example Problem}
	\label{s:example problem}
	This section presents an example problem to demonstrate the process of obtaining the FPG and the advantages of doing so. 
	For this example the algorithm from Sec.~\ref{s:building graphs}.\ref{ss:obtaining FPG} was implemented in the NetworkX package of the Python programming language. 
	While any language with the ability to manipulate a graph should be applicable, NetworkX was chosen due to its built--in features, such as cycle detection and shortest path algorithms.

	The example task is the conceptual sizing of a single--aisle subsonic transport using an MDAO framework with objectives for performance and total weight for a required mission. 
	The full set of analysis codes available for use is given in Table \ref{t:analysis codes}.
	\begin{table}[htbp]
	  \centering
	  \caption{Analysis codes for subsonic transport sizing}
		\begin{tabular}{cc}
		\toprule
		analysis code & description \\
		\midrule
		VSP   & parametric vehicle geometry \\
		PDCYL & wing and fuselage weight estimation \\
		NPSS  & engine sizing and performance \\
		VORLAX & aerodynamics using the vortex lattice method \\
		PMARC & aerodynamics using a low-order panel method \\
		WATE  & engine weight estimation \\
		FLOPSa & mission performance, \\
		  & engine sizing, and weight estimation \\
		FLOPSb & mission performance only \\
		\bottomrule
		\end{tabular}%
	  \label{t:analysis codes}%
	\end{table}%
	Each of these analysis codes contribute individual disciplinary analysis but they are not mutually exclusive. 
	For example, VORLAX and PMARC are both aerodynamics codes that predict inviscid drag but with different levels of fidelity. 
	The Flight Optimization System (FLOPS) is included twice to represent two different uses. 
	FLOPSa denotes FLOPS being executed for both mission performance and engine analysis, while FLOPSb indicates that FLOPS is executed for only mission performance analysis. 
	This representation is useful for ``supercodes'' that are capable of being used in different ways and with different sets of inputs and outputs.

	In this example, the MCG is formed using a consistant variable naming convention and then connecting every variable with the same name, though this is not the only way to form an MCG. 
	Table \ref{t:ins and outs} presents the full list of variables in the leftmost column and then indicates whether the variable is an input or an output of each analysis code. 
	Some of the variables, such as geometry and performance, represent many different variables which have been bundled together due to there similarity. 
	This bundling is not necessary and does not limit the generality of this example; it is done to simplify the presentation.
	\begin{table}[htb!]
	  \centering
	  \caption{Analysis code input and output description}
		\begin{tabular}{ccccccccc}
		\toprule
		variable & \multicolumn{8}{c}{analysis code} \\
		\midrule
			  & VSP   & PDCYL & NPSS  & VORLAX & PMARC & WATE  & FLOPSa & FLOPSb \\
		geometry & in    & in    &       & in    & in    & in    & in    & in \\
		number of engines &       &       & in    &       &       &       & in    & in \\
		mission &       &       &       &       &       &       & in    & in \\
		fuselage weight &       & out   &       &       &       &       & in    & in \\
		wing weight &       & out   &       &       &       &       & in    & in \\
		engine weight &       &       &       &       &       & out   & out   & in \\
		wetted area & out   &       &       &       &       &       & in    & in \\
		inviscid drag &       &       &       & out   & out   &       &       & in \\
		drag  &       &       & in    &       &       &       & out   & out \\
		engine performance &       &       & out   &       &       & in    & out   & in \\
		performance &       &       &       &       &       &       & out   & out \\
		total weight &       & in    &       &       &       &       & out   & out \\
		\bottomrule
		\end{tabular}
	  \label{t:ins and outs}
	\end{table}

	The MCG $M$ is formed in four steps:
	\begin{enumerate}
	\item An analysis block is created for each analysis code using the information in each column of Table \ref{t:ins and outs}. 
	Each analysis block is formed by creating variable nodes for each input and directing them into a single model node, which is directed into another model node, which is then directed into variable nodes corresponding to each output. 
	A sample analysis block is shown for analysis code FLOPSb in Fig.~\ref{f:FLOPSb analysis block}.
	\begin{figure}[htb!]
	  \begin{center}
		\includegraphics[width=6in]{images/MCG_edit_holes}
	  \end{center}
	  \caption{Sample analysis block for analysis code FLOPSb using Table \ref{t:ins and outs}}
	\label{f:FLOPSb analysis block}
	\end{figure}

	\item Variable nodes are created to represent the objectives for performance and total weight.

@@

	\item Connection edges are created from each variable node to every other variable node representing variables with the same name. 
	The direction is determined by whether the variable node has an edge directed into or out of a model node (local input or local output).
	\end{enumerate}
	The resulting MCG is shown in Fig.~\ref{f:MCG}. 
	Emerging cycles are shown in yellow and indicate potential cycles that may exist when an FPG is formed.
	\begin{figure}[htb!]
	  \begin{center}
		\includegraphics[width=6in]{images/MCG_edit}
	  \end{center}
	  \caption{Maximal connectivity graph for the subsonic transport example problem}
	\label{f:MCG}
	\end{figure}

@@

	As presented in Sec.~\ref{s:building graphs}.\ref{ss:obtaining FPG}, the first step to obtaining an FPG is to detect holes. 
	To start, we set the lower indegree limit for every node to be unity. 
	In this case, it turns out that the variable nodes representing the number of engines is a hole for NPSS, FLOPSa, and FLOPSb. 
	This implies that an FPG cannot be obtained from the given MCG, as shown in Fig.~\ref{f:MCG nengines}. 
	In this figure, the green nodes indicate analysis blocks with a hole, and the cyan nodes indicate analysis blocks that became unused after the invalid analyses were removed ((connections were removed, color edges too)).
	\begin{figure}[htb!]
	  \begin{center}
		\includegraphics[width=.6\textwidth]{images/MCG_nengines}
	  \end{center}
	  \caption{The MCG with one global variable missing. The resulting holes are shown in green and unused analyses shown in cyan.}
	\label{f:MCG nengines}
	\end{figure}
	Therefore, we shall now create another variable node to represent the number of engines as a global input and also create connection edges to supply this new input.

	While Sec.~\ref{s:building graphs}.\ref{ss:obtaining FPG} indicated the process for obtaining an FPG, it did not specify how to decide which edges to keep when resolving a collision, that freedom is left to the specific implementation. 
	The benefit of this approach is that standard graph theory algorithms can be used to make these decisions. 
	The first example of this is to obtain an FPG with the fewest possible number of cycles, which are found using the implementation of Johnson's algorithm \cite{Johnson1975} in the Python package NetworkX. 
	The FPG with the fewest cycles is shown in Fig.~\ref{f:FPG fewest cycles}, which indicates one cycle shown in yellow. 
	A similar alternative would be to minimize the number of analysis blocks involved in cycles, counting multiplicity, or to minimize the length of the longest cycle. Since these graphs will always be relatively small (on the order of hundres of nodes) the 
	computational expense for examining all possible FPGs is negligble.
	\begin{figure}[htb!]
	  \begin{center}
		\includegraphics[width=.6\textwidth]{images/FPG_fewest_cycles}
	  \end{center}
	  \caption{FPG with the fewest number of cycles.}
	\label{f:FPG fewest cycles}
	\end{figure}

	Alternatively, it may be desirable to resolve collisions by choosing the higher fidelity codes, such as NPSS in this case.
	One method to ensure that the edges directed from certain analysis blocks are those chosen to resolve conflicts is to use a ranking system. 
	For this case, consider the rankings for each analysis code given in Table \ref{t:rankings}.
	\begin{table}[htbp]
	  \centering
	  \caption{Example ranking of importance}
		\begin{tabular}{cc}
		\toprule
		analysis code & rank \\
		\midrule
		VSP   & 5 \\
		PDCYL & 5 \\
		NPSS  & 4 \\
		PMARC & 4 \\
		FLOPSb & 4 \\
		VORLAX & 3 \\
		WATE  & 3 \\
		FLOPSa & 2 \\
		\bottomrule
		\end{tabular}%
	  \label{t:rankings}%
	\end{table}% 
	Step two in Sec.~\ref{s:building graphs}.\ref{ss:obtaining FPG} (the resolution of collisions) is now conducted by selecting the edges directed from the analysis blocks with the highest rank for each collision. 
	The resulting FPG is shown in Fig.~\ref{f:FPG highest rank}, and this graph has four cycles, which are enumerated in Table \ref{t:cycles}.
	\begin{figure}[htb!]
	  \begin{center}
		\includegraphics[width=.6\textwidth]{images/FPG_highest_rank}
	  \end{center}
	  \caption{FPG obtained using the ranking system.}
	\label{f:FPG highest rank}
	\end{figure}
	\begin{table}[htbp]
	  \centering
	  \caption{Cycles for the FPG obtained using ranking}
		\begin{tabular}{ccccccc}
		\toprule
		output/input & analysis code & output/input & analysis code & output/input & analysis code & output/input \\
		\midrule
		engine weight & FLOPSb & drag  & NPSS  & engine performance & WATE  & engine weight \\
		engine performance & FLOPSb & drag  & NPSS  & engine performance &       &  \\
		fuselage weight & FLOPSb & total weight & PDCYL & fuselage weight &       &  \\
		wing weight & FLOBSb & total weight & PDCYL & wing weight &       &  \\
		\bottomrule
		\end{tabular}%
	  \label{t:cycles}
	\end{table}%

	Finally, now consider the case where it is desired to have the inviscid drag input into FLOPSb be a multi--fidelity input, meaning that it is sought to use multiple analysis codes to calculate the same variable. 
	This is done simply by setting the upper indegree limit for this node as two and then repreating the (automated) process. 
	In this case, both VORLAX and PMARC are retained, resulting in an FPG that omits only FLOPSa.
%\clearpage
%\newtheorem{name}{Printed output}
%\begin{name}
%asdf
%\end{name}
%\begin{proof}[asdf]
%asdf
%\end{proof}

### 2013-04-19 graph_syntax.tex
one last edit
Dr. German's edits

\section{Requirements for New Graph Syntax}
  \label{s:requirements}
  The ultimate goal of the new graph--based syntax presented here is to be able to 
  fully describe the general structure of an MDAO problem independently of any solution information, 
  while still being able to accomodate the more specific case when a solution 
  strategy is applied. In order to achieve that goal 
  the graph needs to accomodate a number of constructs of MDAO problems: 
  \begin{itemize}
    \item Analysis tools and connections between them
    \item Design variables, objectives, and constraints
    \item Local and global properties
    \item Coupling between analyses
    \item Multi-fidelity analyses
  \end{itemize}

  Beyond those basic constructs, there are also three phases of a design process that 
  all need to be representable with the new syntax. Firstly there is the initial problem definition
  phase where the specific analysis tools and design goals are identified. At the end of this phase, 
  a single formal problem formulation is selected specifying design variables, constraints, objectives, 
  analysis tools, etc. Lastly some specific procedure for solving the problem is selected, for example 
  picking an MDAO optimization architecture. Using the proposed graph syntax, each of these phases 
  can be represented with the following three graphs:
  \begin{itemize}
    \item Maximal Connectivity Graph (MCG)
    \item Fundamental Problem Graph (FPG)
    \item Problem Solution Graph (SPG)
  \end{itemize}

  The \emph{maximal connectivity graph} represents the first phase with all 
  analysis codes being considered and all possible connections between them also present. The second graph 
  is the \emph{fundamental problem graph}, which is the smallest possible graph 
  that still fully defines a given problem formulation. Finally, a \emph{specific problem formulation} 
  may be represented by including additional edges and nodes to represent the 
  solution strategy being employed to solve the problem.

  The relationship between these three graphs is depicted in Figs.~\ref{f:tree} and \ref{f:hourglass}. 
  The tree diagram demonstrates the fact that it is generally possible to obtain 
  multiple FPGs from a single maximal connectivity graph. This  may correspond to 
  different down--selections of analysis codes, different connections between them, 
  or both. Each down--selection shrinks the number of possible FPGs that could be reached 
  until only one is remaining. Then, from a single FPG, different PSGs may be obtained by implementing 
  different solution strategies. If you consider the size of a graph to be the sum of all of its
  edges and nodes then the hourglass shape in Fig. \ref{f:hourglass} illustrates how
  the MCG gets reduced to a single FPG, then multiple possible PSGs exist to solve the problem.
  In other words the FPG is obtained from the MCG by removing nodes and edges, 
  and the PSG is obtained from the FPG by adding nodes and edges.

@@

\section{Syntax Definition}
  \label{s:syntax definition}
  In this section we present the necessary graph theory fundamentals to 
  construct the graphs discussed above. 
  The notation used in this work is adapted from Diestel \cite{Diestel2010}. 
  A \emph{graph} is a pair $G = (V,E)$ of sets such that $E \subseteq V \times V$, 
  which means that the elements of $E$ are 2--element subsets of $V$. The set $V$ 
  contains the \emph{vertices} or \emph{nodes} and the set $E$ contains the \emph{edges}.
  For a \emph{directed graph*} (or \emph{digraph}) we construct $E$ as a set of ordered pairs instead 
  of a set of sets. Each ordered pair represents an edge starting at the node 
  indicated by the first entry and directed to the node indicated by the second 
  entry. Edge $e$ = $(x,y)$ may be referred to simply as $xy$. For node $v \in V$ 
  the edges directed out are given by $E(v)$ and the edges directed into $v$ are given 
  by $E^{-1}(v)$; $E(E(v))$ is denoted as $E^2(v)$, and likewise for additional levels. 
  If $E$ is not a one--to--one mapping, $E(v)$ may be the empty set, a single node, or a set of nodes.
  \begin{figure}[htb!]
    \begin{center}
    \includegraphics[width=1.5in]{images/example_directed_graph}
    \end{center}
    \vspace{-20pt}
  \caption{Example directed graph.}
  \label{f:example directed graph}
  \end{figure}
  As an example, for the directed graph shown in Fig.~\ref{f:example directed graph} we have
  \begin{IEEEeqnarray*}{rCl}
  V & = & \{1,2,3,4,5\}, \\
  E & = & \big\{(1,2),(3,2),(1,3),(3,4)\big\}.
  \end{IEEEeqnarray*}

@@

  %does this belong in the analysis block section???? 
  Let $I$ be a nonempty set of counting numbers such that for each $i \in I$ there is a corresponding set $A_i$. 
  The set of sets $\mathcal A = \{A_i \st i \in I\}$ is called an indexed family of sets with index $i$ and 
  indexing set $I$\cite{smith2006}. 
  The union over this family of sets can be described in a few different ways:
  \begin{equation}
  \bigcup_{i \in I} A_i = \bigcup_{A \in \mathcal A} A = \{x \st x \in A \txt{ for some } A \in \mathcal A\}.
  \end{equation}

@@

  \begin{description}
  \item[connection edge:] Exchange of information between two nodes. These edges 
  can either be fixed (can not be removed from graph) or free (can be removed). 
  \item [driven edge:] Passing of information from a driver node to a 
    variable node. A single variable node can have many incomming driven edges. These edges are 
    free and can be added or removed from the graph. 
  \end{description}

@@

  In Alexandrov and Lewis's REMS syntax only two node types were present, variable 
  and function, and only one edge type\cite{alexandrov2004}. We've chosen to rename the ``function'' node 
  type to ``model'' to be more consistent with modern MDAO frameworks. The present work 
  adds one more node types and one more edge type to the syntax to allow descriptive
  graphs for all three phases of the design process. In addition, we introduce the 
  concept of fixed vs. free connection edges to represent the
  static relationship between variable and model nodes from a single analysis, 
  vs the flexible relationship between variables from different analyses.

\subsection{Rules for Nodes and Edges}
  \label{ss:rules}
  There are specific rules for the usage of these nodes and edges.
  The driver node and the driven edge are only allowed to be present in PSGs. All 
  other node and edge types can be present in any of the three graphs, subject to these restrictions: 
  \begin{enumerate}
  \item A model node can only have one edge directed to or from another model node.
  \item A model node can only have fixed connection edges directed in or out.
  \item A model node must have at least one edge directed in and at least one edge 
    directed out.
  \item If a variable node has an outgoing edge to a model node then it may not have 
    any other outgoing edges.
  \end{enumerate}

\subsection{indegree and outdegree}
  \label{s:indegree-outdegree}
  The \emph{indegree} of a node is the number of connection edges directed in and 
  is denoted as $\txt{deg}^-(v)$, and the \emph{outdegree} 
  is the number of connection edges directed out and it is denoted as $\txt{deg}^+(v)$.
  The degree of a given node is only a function of the connection edges 
  attached to it. The number of driven edges is not relevant since any number 
  of drivers could be involved, at different parts of an solution process. For 
  example in a sequential optimization where a global optimizer is first applied
  and a gradient based one is applied second, design variables would have incomming 
  driven edges from both optimizers but this would not affect the indegree of those
  variable nodes.

  We also define the \emph{upper indegree limit}

  \begin{equation}
  \txt{deg}_u^-(v):V \to \mathbb{N}
  \end{equation} 
  and the \emph{lower indegree limit} as
  \begin{equation}
  \txt{deg}_l^-(v):V \to \mathbb{N}.
  \end{equation}
  These user-specified limits govern the number of connection edges that may 
  be directed into a variable node for a valid graph. Consider a variable 
  node $v$ with $\txt{deg}_u^-(v) = \txt{deg}_l^-(v) = 1$. In this case, $v$ 
  must have exactly one incomming explicit edge or the graph is invalid. 
  Two possiblities exist for voilating these limit conditions:

  \begin{description}
    \item[hole: ] The number of incomming edges is less than the lower indegree limit:
      \begin{equation} \txt{deg}^-(v) < \txt{deg}_l^-(v) \label{e:hole} \end{equation}
      This represents a lack of information being supplied to the node.
    \item[collision: ] The number of incomming edges is greater than $ \txt{deg}_u^-(v)$. 
      \begin{equation} \txt{deg}^-(v) > \txt{deg}_u^-(v) \label{e:collision}\end{equation}
  \end{description}

  An algorithm for detecting and managing these violations if proposed in Sec.~\ref{s:building graphs}.\ref{ss:obtaining FPG}.

>>>>

\section{Requirements for New Graph Syntax}
  \label{s:requirements}
  The goal of the new graph--based syntax presented here is to enable the general structure of an MDAO problem to be described independently of any solution information, 
  while still being able to accomodate the more specific case when a solution 
  strategy is applied. In order to achieve this goal, 
  the graph syntax needs to accomodate a number of MDAO problem constructs: 
  \begin{itemize}
    \item Analysis tools and their interconnections
    \item Design variables, objectives, and constraints
    \item Local and global information
    \item Coupling between analyses
    \item Multi-fidelity analyses
  \end{itemize}

The new syntax is intended to represent three phases of the design problem formulation process. 
In the initial problem definition
  phase, the specific analysis tools and design goals are identified. Next, 
  a single formal problem formulation is identified that specifies design variables, constraints, objectives, 
  analysis tools, and all other elements required to represent the overall MDAO problem. 
Lastly, a specific procedure for solving the problem is selected, e.g. selecting an MDAO optimization architecture. Using the proposed graph syntax, these phases 
  can be represented with the following three graphs:
  \begin{itemize}
    \item Maximal Connectivity Graph (MCG)
    \item Fundamental Problem Graph (FPG)
    \item Problem Solution Graph (PSG)
  \end{itemize}

  The \emph{maximal connectivity graph} represents the first phase of the problem formulation with all 
  analysis codes being considered and all possible connections between them also present. The second graph 
  is the \emph{fundamental problem graph}, which is the smallest possible graph 
  that still fully defines a given problem formulation that could in principle be solved by an MDAO technique. Finally, a \emph{specific problem formulation} 
  may be represented by including additional edges and nodes to represent the 
  solution strategy being employed to solve the problem.

  The relationship between these three graphs is depicted in Figs.~\ref{f:tree} and \ref{f:hourglass}. 
  The tree diagram demonstrates the fact that it is generally possible to obtain 
  multiple FPGs from a single maximal connectivity graph. This  may correspond to 
  different down--selections of analysis codes, different connections between them, 
  or both. Each down--selection shrinks the number of possible FPGs that could be reached 
  until only one is remaining. Then, from a single FPG, different PSGs may be obtained by implementing 
  different solution strategies. If you consider the size of a graph to be the sum of all of its
  edges and nodes then the hourglass shape in Fig. \ref{f:hourglass} illustrates how
the FPG is obtained from the MCG by removing nodes and edges, 
  and the PSG is obtained from the FPG by adding nodes and edges.

@@

\section{Syntax Definition}
  \label{s:syntax definition}
  In this section, we present the necessary graph theoretic fundamentals to 
  construct the graphs discussed in Sec.~\ref{s:requirements}. 
  The notation used in this work is adapted from Diestel \cite{Diestel2010}. 
  A \emph{graph} is a pair $G = (V,E)$ of sets such that $E \subseteq V \times V$, 
  which means that the elements of $E$ are 2--element subsets of $V$. The set $V$ 
  contains the \emph{vertices} or \emph{nodes} and the set $E$ contains the \emph{edges}.
  For a \emph{directed graph*}((DJP: why the asterisk?)) (or \emph{digraph}) we construct $E$ as a set of ordered pairs instead 
  of a set of sets. Each ordered pair represents an edge starting at the node 
  indicated by the first entry and directed to the node indicated by the second 
  entry. Edge $e$ = $(x,y)$ may be referred to simply as $xy$. For node $v \in V$ 
  the edges directed out from $v$ are denoted by $E(v)$ and the edges directed into $v$ are given 
  by $E^{-1}(v)$. 
  If $E$ is not a one--to--one mapping, $E(v)$ may be the empty set, a single node, or a set of nodes.
  \begin{figure}[htb!]
    \begin{center}
    \includegraphics[width=1.5in]{images/example_directed_graph}
    \end{center}
    \vspace{-20pt}
  \caption{Example directed graph.}
  \label{f:example directed graph}
  \end{figure}
  As an example, for the directed graph shown in Fig.~\ref{f:example directed graph} we have
  \begin{IEEEeqnarray*}{rCl}
  V & = & \{1,2,3,4,5\}, \\
  E & = & \big\{(2,1),(3,2),(1,3),(3,4)\big\}.
  \end{IEEEeqnarray*}

@@

  %does this belong in the analysis block section???? 
  Let $I$ be a nonempty set such that for each $i \in I$ there is a corresponding set $A_i$. 
  The set of sets $\mathcal A = \{A_i \st i \in I\}$ is called an indexed family of sets with index $i$ and 
  indexing set $I$\cite{smith2006}. 
  The union over this family of sets can be described in a few different ways:
  \begin{equation}
  \bigcup_{i \in I} A_i = \bigcup_{A \in \mathcal A} A = \{x \st x \in A \txt{ for some } A \in \mathcal A\}.
  \end{equation}

@@

  \begin{description}
  \item[connection edge:] Exchange of information between two nodes. These edges 
  can either be fixed (cannot be removed from graph) or free (can be removed). 
  \item [driven edge:] Passing of information from a driver node to a 
    variable node. A single variable node can have many incomming driven edges. These edges are 
    free and can be added or removed from the graph. 
  \end{description}

@@

  In Alexandrov and Lewis's REMS syntax only two node types (variable and function) and one edge type are included \cite{alexandrov2004}. 
 We have chosen to adopt the terminology ``model'' node instead of ``function'' node to be consistent with modern MDAO framework terminonlogy. The present work 
  adds one additional node types and one additional edge type to the syntax to allow descriptive
  graphs for all three phases of the design problem formulation process. In addition, we introduce the 
  concept of fixed vs. free connection edges to represent the
  static relationship between variable and model nodes from a single analysis, 
  vs the flexible relationship between variables from different analyses.

\subsection{Rules for Nodes and Edges}
  \label{ss:rules}
  A rule set is provided for the usage of these nodes and edges.
  The driver node and the driven edge are allowed to be present only in PSGs.
  Other node and edge types can be present in any of the three graph types, subject to the following restrictions: 
  \begin{enumerate}
  \item A model node may have only one edge directed to or from another model node.
  \item A model node may have only fixed connection edges directed in or out.
  \item A model node must have at least one edge directed in and at least one edge 
    directed out.
  \item If a variable node has an outgoing edge to a model node, then it may not have other outgoing edges.
  \end{enumerate}

\subsection{indegree and outdegree}
  \label{s:indegree-outdegree}
  The \emph{indegree} of a node is the number of connection edges directed in and 
  is denoted as $\txt{deg}^-(v)$, and the \emph{outdegree} 
  is the number of connection edges directed out and it is denoted as $\txt{deg}^+(v)$.
  The degree of a given node is a function only of the connection edges 
  attached to it. The number of driven edges is not relevant because any number 
  of drivers could be involved in different parts of an solution process. For 
  example, in a sequential optimization in which a global optimizer is applied first ((BG: first and second?))
  and a gradient based optimizer is applied second, design variables would have incomming 
  driven edges from both optimizers but this would not affect the indegree of those
  variable nodes.

  Specifially for variable nodes we also define the \emph{upper indegree limit}

  \begin{equation}
  \txt{deg}_u^-(v):V \to \mathbb{N}
  \end{equation} 
  and the \emph{lower indegree limit} as
  \begin{equation}
  \txt{deg}_l^-(v):V \to \mathbb{N}.
  \end{equation}
  These user-specified limits govern the number of connection edges that may 
  be directed into a variable node for a valid graph. Consider a variable 
  node $v$ with $\txt{deg}_u^-(v) = \txt{deg}_l^-(v) = 1$. In this case, $v$ 
  must have exactly one incomming explicit edge or the graph is deemed invalid. 
  Two possiblities exist for voilating these limit conditions:

  \begin{description}
    \item[hole: ] The number of incomming edges is less than the lower indegree limit:
      \begin{equation} \txt{deg}^-(v) < \txt{deg}_l^-(v) \label{e:hole} \end{equation}
      A hole represents a lack of information being supplied to the node.
    \item[collision: ] The number of incomming edges is greater than $ \txt{deg}_u^-(v)$. 
      \begin{equation} \txt{deg}^-(v) > \txt{deg}_u^-(v) \label{e:collision}\end{equation}
  \end{description}

  The presence of holes and collisions in a graph represent issues that will give
  rise to an invalid problem formulation. An algorithm for detecting and 
  managing these violations if proposed in Sec.~\ref{s:building graphs}.\ref{ss:obtaining FPG}.
